### 2015-08-17 doc/thesis/conclusions.tex
Improving the informal description about table tags

After designing and implementing Typed Lua, we performed several
case studies to evaluate how successful we were in our goal of
providing an usable type system.
We evaluated 29 modules from 8 different case studies,
and we could give precise static types to 83\% of the 449
members that these modules export.
In the median, we could give precise static types to 89\%
of the members from each module.
Our evaluation results showed that our type system can statically
type check several Lua idioms and features, though the evaluation
results also exposed several limitations of our type system.
We found that the three main limitations of our type system are
the lack of intersection types, parametric polymorphism, and operator overloading.
Overcoming these limitations is our major target for future work,
as it will allow us to statically type check more programs.

@@

Finally, we believe that Typed Lua is a major contribution to the Lua community,
because it offers a framework that programmers can use to document,
test, and better structure their applications.
For libraries where a full conversion to static type checking should
prove unfeasible or too much work, the community can use Typed Lua
just to document the external interfaces of the libraries,
giving the benefits of static type checking to the users of these
libraries.
In fact, we already have user feedback from Lua programmers that are
using Typed Lua in their projects.
For instance, ZeroBrane Studio is an IDE for Lua development that is
starting to use Typed Lua to perform static analysis in Lua code.

>>>>

After designing and implementing Typed Lua, we performed several
case studies to evaluate how successful we were in our goal of
providing an usable type system.
We evaluated 29 modules from 8 different case studies,
and we could give precise static types to 83\% of the 449
members that these modules export.
For half of the modules, we could give precise static types to
at least 89\% of the members from each module.
Our evaluation results showed that our type system can statically
type check several Lua idioms and features, though the evaluation
results also exposed several limitations of our type system.
We found that the three main limitations of our type system are
the lack of intersection types, parametric polymorphism, and operator overloading.
Overcoming these limitations is our major target for future work,
as it will allow us to statically type check more programs.

@@

Finally, we believe that Typed Lua is a major contribution to the Lua community,
because it offers a framework that programmers can use to document,
test, and better structure their applications.
For libraries where a full conversion to static type checking should
prove unfeasible or too much work, the community can use Typed Lua
just to document the external interfaces of the libraries,
giving the benefits of static type checking to the users of these
libraries.
In fact, we already have user feedback from Lua programmers that are
using Typed Lua in their projects.
For instance, ZeroBrane Studio is an IDE for Lua development that is
evaluating the use of Typed Lua to perform static analysis in Lua code.

### 2015-08-17 doc/thesis/typedlua.tex
Improving the informal description about table tags

Typed Lua tags table types as either \emph{unique}, \emph{open},
\emph{fixed}, or \emph{closed} to identify which operations are safe.
The table constructor describes \emph{unique} table types,
as they are tables that we know all the fields and that there is
no alias to it.
Any alias to \emph{unique} tables makes them \emph{open},
as we use this tag to keep track of \emph{unique} aliasing.
In the previous example, \texttt{person} has an \emph{unique} type
because it has no alias.
In Section \ref{sec:oop} we will show that we use the tag
\emph{fixed} to describe the type of classes.
Type annotations define \emph{closed} table types, as this tag
describes table types that do not provide any guarantees about keys
that do not appear in the table type.
Typed Lua also has different subtyping rules that handle these
different tags, which we explain in Section \ref{sec:subtyping}.

Typed Lua uses three rules to decide whether it is sound to
allow field assignment to change the type of \emph{unique} and
\emph{open} tables:
the table must assigned to a local variable to the current block,
the new type must only add new fields, and
the variable cannot have been assigned in another function.

@@

Back to the rules that Typed Lua uses to allow the refinement
of table types, the following example shows that it is not sound
to allow the type of the fields to change:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1942           --> compile-time warning
    person.lastname = 2013            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, both third and fourth lines are unsound because
they are trying to change the type of fields that already exist.

It is also not sound to allow changing the type of an alias:
\begin{verbatim}
    local person = {}
    local bogus = person
    person.firstname = "Lou"
    person.lastname = "Reed"
    bogus.firstname = true            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

@@

Typed Lua uses \emph{closed} table types along with the type
\texttt{self} to represent objects.
We use the following example to discuss this feature:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

>>>>

Typed Lua tags table types as either \emph{unique}, \emph{open},
\emph{fixed}, or \emph{closed} to identify which operations are safe.
Type annotations define \emph{closed} table types, as this tag
describes table types that do not provide any guarantees about keys
that do not appear in the table type.
In contrast, \emph{unique}, \emph{open}, and \emph{fixed} table types
guarantee that a table does not contain any keys which are
not listed in its table type, though they provide different guarantees
over their references.
An \emph{unique} table type guarantees that it has only one reference.
In particular, this is the case of table constructors, so they
always have \emph{unique} table types.
In the previous example, \texttt{person} has an \emph{unique} type
because it has no alias.
Any alias to \emph{unique} tables makes them \emph{open},
as we use this tag to keep track of \emph{unique} aliasing.
This means that \emph{open} table types can only have \emph{closed}
aliases, and it is guaranteed that only one reference remains \emph{open}.
Finally, \emph{fixed} table types can have any number of \emph{closed}
and \emph{fixed} aliases, so its type cannot change.
In Section \ref{sec:oop} we will show that we use \emph{fixed} table
types to describe the type of classes.
Typed Lua also has different subtyping rules that handle these
different tags, which we explain in Section \ref{sec:subtyping}.

Typed Lua uses three rules to check whether field assignment can
change the type of \emph{unique} and \emph{open} tables:
the table must be assigned to a local variable to the current block,
the new type must only add new fields, and
the variable cannot have been assigned in another function.

@@

Back to the rules that Typed Lua uses to allow the refinement
of table types, the following example shows that it is forbidden
to change the type of existing fields:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1942           --> compile-time warning
    person.lastname = 2013            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, both third and fourth lines are not valid because
they are trying to change the type of fields that already exist.
Even though it would be sound to allow changing the type of fields
in \emph{unique} table types because they cannot have any alias,
this could still lead to a weak static type checking approach.
In other words, allowing the type of the fields to evolve from a
base type to an union type or even to the dynamic type could hide
some type errors, so our design choice was to keep it as precise
as we could.
Moreover, it would be unsound to allow changing the type of fields
in \emph{open} table types because they can have \emph{closed} aliases,
so it made more sense to have the same rule to both table types.

The following example shows that it is not allowed to change
the type of an alias:
\begin{verbatim}
    local person = {}
    local bogus = person
    person.firstname = "Lou"
    person.lastname = "Reed"
    bogus.firstname = true            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

@@

Typed Lua uses \emph{closed} table types along with the type
\texttt{self} to represent objects.
We use the following example to discuss this feature:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }

    function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}
